### 2019-07-06 1-Preamble/5-Abstract/abstract.tex
Not sure if I say enough about the lower-bounds...

%
% \TODOL{Raccourcir juste un peu ?}
% ----------------------------------------------------------------------------
% \begin{small} % XXX
% Chapitre 1
Dans cette thèse de doctorat, nous étudions les réseaux sans fil et les appareils reconfigurables qui peuvent accéder à des réseaux de type radio intelligente, dans des bandes non licenciées et sans supervision centrale.
Nous considérons notamment des réseaux actuels ou futures de l'Internet des Objets (IoT), avec l'objectif d'augmenter la durée de vie de la batterie des appareils, en les équipant d'algorithmes d'apprentissage machine peu coûteux mais efficaces, qui leur permettent d'améliorer automatiquement l'efficacité de leurs communications sans fil.
%  (Chapitre~\ref{chapter:1}).
% Chapitre 4
Nous proposons deux modèles de réseaux IoT, et nous montrons empiriquement, par des simulations numériques et une validation expérimentale réaliste, le gain que peuvent apporter nos méthodes, qui se reposent sur l'apprentissage par renforcement.
%  (Chapitre~\ref{chapitre:4}).
% Chapitre 2
Les différents problèmes d'accès au réseau sont modélisés avec des Bandits Multi-Bras (MAB), mais l'analyse de la convergence d'un grand nombre d'appareils jouant à un jeu collaboratif sans communication ni aucune coordination reste délicate,
lorsque les appareils suivent tous un modèle d'activation aléatoire.
% En effet, même le nombre de périphériques actifs à chaque instant est hautement imprévisible, rendant l'environnement non pas stationnaire et évoluant rapidement.
% Chapitres 5 et 6
Le reste de ce manuscrit étudie donc deux modèles restreints, d'abord des bandits multi-joueurs dans des problèmes stationnaires, puis des bandits mono-joueur non stationnaires.
%  (Chapitre~\ref{chapitre:5})
%  (Chapitre~\ref{chapitre:6})
% Chapitre 3
Nous détaillons également une autre contribution, la bibliothèque Python open-source SMPyBandits, qui permet des simulations numériques de problèmes MAB, qui couvre les modèles étudiés et d'autres.
% (Chapitre~\ref{chapter:3})
% \end{small} % XXX

>>>>

%
% \TODOL{Raccourcir juste un peu ?}
% ----------------------------------------------------------------------------
% \begin{small} % XXX
% Chapitre 1
\noindent
Dans cette thèse de doctorat, nous étudions les réseaux sans fil et les appareils reconfigurables qui peuvent accéder à des réseaux de type radio intelligente, dans des bandes non licenciées et sans supervision centrale.
Nous considérons notamment des réseaux actuels ou futures de l'Internet des Objets (IoT), avec l'objectif d'augmenter la durée de vie de la batterie des appareils, en les équipant d'algorithmes d'apprentissage machine peu coûteux mais efficaces, qui leur permettent d'améliorer automatiquement l'efficacité de leurs communications sans fil.
%  (Chapitre~\ref{chapter:1}).
% Chapitre 4
Nous proposons deux modèles de réseaux IoT, et nous montrons empiriquement, par des simulations numériques et une validation expérimentale réaliste, le gain que peuvent apporter nos méthodes, qui se reposent sur l'apprentissage par renforcement.
%  (Chapitre~\ref{chapitre:4}).
% Chapitre 2
Les différents problèmes d'accès au réseau sont modélisés avec des Bandits Multi-Bras (MAB), mais l'analyse de la convergence d'un grand nombre d'appareils jouant à un jeu collaboratif sans communication ni aucune coordination reste délicate,
lorsque les appareils suivent tous un modèle d'activation aléatoire.
% En effet, même le nombre de périphériques actifs à chaque instant est hautement imprévisible, rendant l'environnement non pas stationnaire et évoluant rapidement.
% Chapitres 5 et 6
Le reste de ce manuscrit étudie donc deux modèles restreints, d'abord des bandits multi-joueurs dans des problèmes stationnaires, puis des bandits mono-joueur non stationnaires.
%  (Chapitre~\ref{chapitre:5})
%  (Chapitre~\ref{chapitre:6})
% Chapitre 3
Nous détaillons également une autre contribution, la bibliothèque Python open-source SMPyBandits, qui permet des simulations numériques de problèmes MAB, qui couvre les modèles étudiés et d'autres.
% (Chapitre~\ref{chapter:3})
% \end{small} % XXX

### 2019-07-06 1-Preamble/5-Abstract/resume_fr.tex
Not sure if I say enough about the lower-bounds...

    \item
    La dernière contribution de cette thèse est notre article \cite{Besson2018DoublingTricks}.
    % , résumé en Annexe~\ref{app:2:DoublingTricks}.
    Cet article donne une revue de la littérature sur l'utilisation de la technique de doublements successifs de l'horizon (``doubling trick'') pour les problèmes de bandits,
    ainsi qu'une analyse unifiée et plus générique de deux familles de doublements.
\end{itemize}

>>>>

    % \item
    % La dernière contribution de cette thèse est notre article \cite{Besson2018DoublingTricks}.
    % % , résumé en Annexe~\ref{app:2:DoublingTricks}.
    % Cet article donne une revue de la littérature sur l'utilisation de la technique de doublements successifs de l'horizon (``doubling trick'') pour les problèmes de bandits,
    % ainsi qu'une analyse unifiée et plus générique de deux familles de doublements.
\end{itemize}

### 2019-07-06 2-Chapters/1-Chapter/chapter1.tex
Not sure if I say enough about the lower-bounds...

\paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single piece of software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we produced a Python library of simulation of MAB problems.
%
We wrote the most exhaustive open-source simulation library for MAB problems, called SMPyBandits, which is published online under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits} and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations discussed in the rest of this thesis.

@@

    \item
    We developed SMPyBandits, the most exhaustive open-source simulation library for MAB problems, that we published on-line under an open-source licence (MIT) \cite{SMPyBanditsJMLR,SMPyBandits}.
    We present it in details in Chapter~\ref{chapter:3}.

>>>>

\paragraph{Chapter~\ref{chapter:3}.}
%
Our exploration of the large number of MAB algorithms and models developed in the recent literature
has given us the ambition to write a single piece of software allowing anyone to easily implement new models and algorithms, in order to compare the existing ones and empirically explore the performances of newly proposed algorithms.
To tackle this goal, we produced a Python library of simulation of MAB problems.
%
To the best of our knowledge, we wrote the most exhaustive open-source library for bandits. Our library SMPyBandits is published online, under an open-source licence \cite{SMPyBanditsJMLR,SMPyBandits}, and hosted on \href{https://GitHub.com/SMPyBandits}{\texttt{GitHub.com/SMPyBandits}}.
We present in details its architecture and its features in Chapter~\ref{chapter:3}, along with different examples of its usage.
A full documentation is available online, as well as exhaustive instructions to reproduce the simulations discussed in the rest of this thesis.

@@

    \item
    We developed SMPyBandits, an exhaustive open-source simulation library for MAB problems, that we published on-line under an open-source licence (MIT) \cite{SMPyBanditsJMLR,SMPyBandits}.
    We present it in details in Chapter~\ref{chapter:3}.

### 2019-07-06 2-Chapters/2-Chapter/chapter2.tex
Not sure if I say enough about the lower-bounds...

    % \item
    MAB can also be applied to a broader setting of \textbf{online content recommandation}.
    % with two options (A/B testing), or more than two options.
    The seminal work of \cite{Li10} studies the application of contextual bandit to online advertisement
    or news articles recommendation, as it is used in practice on platforms such as Microsoft's Bing news website,
    or in video streaming platform like YouTube or Netflix.
    In such models, the arms correspond to items to recommend (\eg, articles or movies), and the contexts contain features about each user of the system.
    An interesting work is \cite{Louedec16}, who studies slowly-varying non-stationary models applied to recommender systems.

@@

\paragraph{Problem-dependent lower-bound in $\Omega(\log T)$.}
%
The following two theorems were proven in \cite{LaiRobbins85}.
These lower-bounds are of highest interest to design efficient algorithms,
and a significant part of the research on MAB algorithms has focused on finding algorithms that matches the Lai and Robbins' lower-bound asymptotically.
This means that a finite-time or an asymptotic upper-bound is proven on the regret of algorithm $\cA$, that asymptotically matches the lower-bound, with the same constant in the big-$\cO$ notation (in which case we say that the algorithm is asymptotically \emph{optimal}), or with a larger constant (the algorithm is said to be \emph{order-optimal} in such case).

@@

Theorem~2.12 from \cite{Slivkins2019} specifies a possible value for the constant $C_I$.

@@

Moreover, it is interesting to note that the first lower-bound of Theorem~\ref{thm:2:forSecondLogTLowerBound} can be directly used to design efficient algorithms\footnote{~Tracking this quantity, by using empirical estimates of the means, is used for instance in the OSSB algorithm proposed in \cite{Combes17}, which is proven to attain the lower-bound for a wider range of problems (for problems called structured stochastic bandits).}, as thanks to the regret decomposition given in Proposition~\ref{prop:2:RegretDecomposition} above, the expression of $C_I$ essentially says that any efficient algorithm
should sample each sub-optimal arm $k$ about $\log(T)/\kl(\mu_k,\mu^*)$ times in the total of $T$ time steps.
%
Many algorithms have been proven to achieve logarithmic regret in the stochastic case,
and in particular it is the case of the algorithms used in this thesis, $\UCB_1$ from \cite{Auer02}, Thompson sampling from \cite{Thompson33} and analyzed in \cite{AgrawalGoyal11,Kaufmann12Thompson}, and \klUCB{} from \cite{Garivier11KL,KLUCBJournal}.
%
Such bounds are valid in different settings, and in particular $\UCB_1$ is order-optimal for bounded rewards or one-dimensional exponential families,
while Thompson sampling is optimal for bounded rewards, and \klUCB{} have been proven to be optimal for both cases.
%
Chapter~16 of \cite{LattimoreBanditAlgorithmsBook} contains more details.

@@

An extension of this strategy is to consider not a fixed time $T_0$ but a random time $\tau$ at which exploration stops.
This time $\tau$ must be a \emph{stopping time}\footnote{See Chapter~3 of \cite{LattimoreBanditAlgorithmsBook}, and we use this notion again in Section~\ref{sec:6:GLRklUCB_Algorithm}.},
in the sense that it is a measurable random variable, dependent of the past observations. The strategy is then referred to as ``explore-then-commit'' (ETC), and the idea is to use a statistical test at every time step $t$, and stop as soon as enough samples were collected to effectively identify the best arm with a certain confidence level $\delta$.
Choosing $\delta \propto 1/T$ and using a lower-bound $d$ on the gaps typically lead to an order-optimal algorithm as shown in \cite{GarivierETC2016}.

@@

We present the regret bound, which was first stated in \cite{Auer02}, and to explain how Hoeffding's inequality is used we prove it below.
Note that for the rest of the manuscript, we use the \klUCB{} algorithm (and more subtle concentration inequality) whenever we analyze the regret of a newly introduced algorithm (\ie, in Chapters~\ref{chapter:5} and \ref{chapter:6}),
but we think the simpler proof of $\UCB_1$ is enlightening for the reader.
% but we preferred to start with a simpler proof in the single-player stationary case.
%
This theorem shows that $\UCB_1$ achieves an order-optimal problem-dependent regret, bounded by $\cO(K\log(T)/\Delta^2)$ (if $\Delta \eqdef \min\limits_{k: \Delta_k>0} \Delta$), and the result also has the advantage of being non asymptotic.

@@

We see in this Theorem~\ref{thm:2:UCBregretBound} that $\UCB_1$ has the advantage of not relying on any prior knowledge of the problem difficulty, of being anytime and computationally simple, for both memory and time aspects, and it achieves an order-optimal problem-dependent regret upper bound in $\cO((\sum_{k: \Delta_k>0} \frac{1}{\Delta_k})\log(T))$.
%
$\UCB_1$ requires a constant storage capacity for each arm (storing $X_k(t)$ and $N_k(t)$ require two integers or a \texttt{float} and an integer, \ie, $\bigO{1}$), giving a storage capacity of $\bigO{K}$, independently of the horizon $T$.
It also needs to perform some mathematical operations, but a square-root and a logarithm are cheap, thus it costs a $\bigO{1}$ constant time for each arm and time step, thus a time $\bigO{K T}$ for the $T$ time steps.

@@

For Bernoulli distributions, \klUCB{} builds tighter confidence intervals than $\UCB_1$, in the sense that $U_k^{\klUCB}(t) < U_k^{\UCB}(t)$
but with the same probability coverage.
This is justified by Pinkser's inequality, which gives $\kl(x,y) \geq 2(x-y)^2$.
%
The \KLUCB{} and \klUCB{} algorithms achieve finite-time regret logarithmic upper-bounds on their regret, and have been proven to be asymptotically optimal, respectively for one-dimensional exponential families and for bounded rewards distributions \cite{Garivier11KL,KLUCBJournal}.
% FIXME

@@

For problems with Bernoulli distributions on the $K$ arms,
the best choice of prior is a Beta distribution, that starts as an initially uniform prior (\ie, $\forall k, \pi_k(0) \eqdef \mathrm{Beta}(1,1)$),
and maintains a distribution $\pi_k(t) \eqdef \mathrm{Beta}(X_k(t)+1,N_k(t)-X_k(t)+1)$,
as $X_k(t)$ counts the number of successes observed on arm $k$, and $N_k(t)-X_k(t)$ counts the number of failures.
%
We detail the Thompson sampling algorithm in this case, below in Algorithm~\ref{algo:2:ThompsonSampling}, as it corresponds to the case used in Chapter~\ref{chapter:4} in Section~\ref{sub:41:numericalResults}.
%
In this setting, as well as any one-dimensional exponential families,
% for Gaussian distributions,
Thompson sampling was proven to achieve asymptotical optimal problem-dependent bounds \cite{Kaufmann12Thompson,AgrawalGoyal11}.

@@

Bayes-UCB from \cite{Kaufmann12BUCB} also uses the Bayesian point of view, updated like for Thompson sampling, but used in a different way.
Instead of sampling from the posteriors and playing the arms with maximum sample, Bayes-UCB computes the $(1-1/(t (\log(t))^5)$ quantile of each arm, at time $t$, and plays the arm with the largest quantile.
It was also proven to achieve asymptotical optimal problem-dependent bounds \cite{Kaufmann12BUCB}.
%
Bayes-UCB is comparable to Thompson sampling in terms of memory complexity, but it can be slower, as computing a quantile is generally more costly than just sampling from a distribution.
% a Beta prior and posterior, initially uniform (\ie, $\mathrm{Beta}(1,1)$) (\texttt{BayesUCB}).
% But in particular for Bernoulli distributed arms and for Beta posteriors and priors, Bayes-UCB is only about twice as slow as Thompson sampling.

@@

\textbf{BESA} (Best Empirical Sampled Average) was introduced in \cite{Baransi2014}, in order to fix a simple observation:
if arm $1$ has $100$ samples and arm $2$ has $50$ samples, it should make more sense to only use $50$ samples of arm $1$ to compute and compare their empirical means.
Based on this idea, the BESA algorithm for $K=2$ arms is quite simple: at each time step, if arms $1$ and $2$ have respectively $n_1$ and $n_2$ samples, and if $n_1>n_2$ (for instance), first it sub-samples $n_2$ observations from $n_1$, then it computes the mean of these observations, and play the arm maximizing this mean (now that there is the same number of observations for both arms, it makes more sense to compare these quantities).
It was analyzed partially, for the two-armed case it was proven to be asymptotically optimal for one-dimensional exponential families, and the authors illustrated that it can be very efficient empirically.
Analyzing it in the generic case was found to be difficult, and it is still an open problem.
% We implemented it using a binary tournament implement in a naive but efficient way (not using recursive functions, \ie, using the \texttt{BESA} class with \texttt{''non\_recursive''=True}).
% By using such binary tournament, the extension $K>2$ arms is costly in terms of both time and memory, as illustrated below, and suffers from an exponential blow-up when $K$ increases.

@@

More recently, the \klUCB-switch from \cite{GarivierHadiji2018} is the first algorithm to be proven to
enjoy simultaneously a distribution-free regret bound of optimal order $\sqrt{KT}$, and a distribution-dependent regret bound of optimal order as well, that is, matching the $C_I \ln(T)$ lower bound by \cite{LaiRobbins85}.
It is an index policy that uses modified versions of the indexes of \klUCB$^+$ for arms that are not sampled much (see \cite{Honda2019} for more details), and of MOSS$^+$ for the other arms.
If the two indexes
% $U^{\KLUCB+}_k(t)$ and $U^{\mathrm{MOSS}+}_k(t)$
are defined by
$U^{\mathrm{MOSS}+}_k(t) \eqdef \frac{X_k(t)}{N_k(t)} + \max\left(0, \sqrt{ \log\left(T / (K N_k(t))\right) / N_k(t)} \right)$
and
$U^{\KLUCB+}_k(t) \eqdef \sup\limits_{q \in [0,1]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \log(T / (K N_k(t))) / N_k(t) \right\}$,
and if $f(T, K) = \lfloor (T/K)^{\gamma}\rfloor$ for $\gamma=1/5$,
then \klUCB-switch uses the index defined as follows:
\begin{align*}
    U^{\klUCB\text{-}\mathrm{switch}}_k(t) & \eqdef \begin{cases}
        U^{\KLUCB+}_k(t) & \text{if } N_k(t) \leq f(T, K), \\
        U^{\mathrm{MOSS}+}_k(t) & \text{if } N_k(t) > f(T, K).
    \end{cases}.
\end{align*}
%

@@

\paragraph{Summary of this review of MAB algorithms.}
%
We gave an overview of the most well known families of algorithms for multi-armed bandits,
focussing on algorithms designed and analyzed for the single-player stationary and stochastic MAB model.
Except the naive and simple strategies given as introductory examples, most of the policies presented above are order-optimal for Bernoulli distributed problems or bounded rewards (or one-dimensional exponential families),
and some of them are known to be optimal in different settings.
%
To clearly understand the difference in their empirical behaviors, we illustrated the performance of two naive and two simple strategies against three efficient ones, on the example of Bernoulli-distributed problem used for the online demonstration from Section~\ref{sec:2:notations} (see Figures~\ref{fig:2:example_of_a_5_arm_bandit_problem} and \ref{fig:2:example_of_a_5_arm_bandit_problem__step100}).
This also serves as a first example of numerical simulation performed with our library SMPyBandits, before the large-scale experiments of Section~\ref{sec:3:reviewSPAlgorithms} below.
%
In the rest of this thesis, we mainly use the $\UCB_1$ or Thompson sampling algorithms when simplicity and realistic deployment is favored (in Chapter~\ref{chapter:4}), and the \klUCB{} algorithm when we give mathematical developments (in Chapters~\ref{chapter:5} and \ref{chapter:6}).

>>>>

    % \item
    MAB can also be applied to a broader setting of \textbf{online content recommandation}.
    % with two options (A/B testing), or more than two options.
    The seminal work of \cite{Li10} studies the application of contextual bandit to news articles recommendation, as it is used in practice on platforms such as Microsoft's Bing news website.
    Other famous examples include online advertisement or in other kinds of content recommendation, such as for video streaming platforms like YouTube or Netflix.
    In such models, the arms correspond to items to recommend (\eg, ads, articles or movies), and the contexts contain features about each user of the system.
    An interesting work is \cite{Louedec16}, who studies slowly-varying non-stationary models applied to recommender systems.

@@

\paragraph{Problem-dependent lower-bound in $\Omega(\log T)$.}
\label{par:2:def_optimal_orderoptimal}
%
The following two theorems were proven in \cite{LaiRobbins85}.
These lower-bounds are of highest interest to design efficient algorithms,
and a significant part of the research on MAB algorithms has focused on finding algorithms that matches the Lai and Robbins' lower-bound asymptotically \cite{LaiRobbins85}.
%
This means that a finite-time or an asymptotic upper-bound is proven on the regret of algorithm $\cA$, that asymptotically matches the lower-bound.
When the regret upper-bound matches the lower-bound with the same constant as in the big-$\cO$ notation, we say that the algorithm is asymptotically \emph{optimal},
otherwise the algorithm is said to be \emph{order-optimal} if it matches the lower-bound with with a larger constant.

@@

Theorem~2.12 from \cite{Slivkins2019} specifies a possible value for the constant $C_I$.
Note that the lower-bound of Lai and Robbins, proved in \cite{LaiRobbins85}, is more general and applies to any one-dimensional exponential family $\cE$, by replacing the Bernoulli Kullback-Leibler divergence $\kl$ with the KL divergence of this family, $\kl_{\cE}$.

@@

It is interesting to note that the first lower-bound of Theorem~\ref{thm:2:forSecondLogTLowerBound} can be directly used to design efficient algorithms\footnote{~Tracking this quantity, by using empirical estimates of the means, is used for instance in the OSSB algorithm proposed in \cite{Combes17}, which is proven to attain the lower-bound for a wider range of problems (for problems called structured stochastic bandits).}, as thanks to the regret decomposition given in Proposition~\ref{prop:2:RegretDecomposition} above, the expression of $C_I$ essentially says that any efficient algorithm
should sample each sub-optimal arm $k$ about $\log(T)/\kl(\mu_k,\mu^*)$ times in the total of $T$ time steps.
%
Many algorithms have been proven to achieve logarithmic regret in the stochastic case,
and in particular it is the case of the algorithms used in the rest of thesis, $\UCB_1$ from \cite{Auer02}, Thompson sampling from \cite{Thompson33} and analyzed in \cite{AgrawalGoyal11,Kaufmann12Thompson}, and \klUCB{} from \cite{Garivier11KL,KLUCBJournal}.
%
Such bounds are valid in different settings:
$\UCB_1$ is order-optimal for bounded rewards or one-dimensional exponential families (\ie, it matches the asymptotic lower-bound of \cite{LaiRobbins85}),
while Thompson sampling is optimal for binary rewards, and \klUCB{} and \KLUCB{} have been proven to be optimal respectively for both cases.
%
Chapter~16 of \cite{LattimoreBanditAlgorithmsBook} contains more details.

@@

An extension of this strategy is to consider not a fixed time $T_0$ but a random time $\tau$ at which exploration stops.
This time $\tau$ must be a \emph{stopping time}\footnote{See Chapter~3 of \cite{LattimoreBanditAlgorithmsBook}, and we use this notion again in Section~\ref{sec:6:GLRklUCB_Algorithm}.},
in the sense that it is a measurable random variable, dependent of the past observations. The strategy is then referred to as ``explore-then-commit'' (ETC), and the idea is to use a statistical test at every time step $t$, and stop as soon as enough samples were collected to effectively identify the best arm with a certain confidence level $\delta$.
Choosing $\delta \propto 1/T$ and using a lower-bound $d$ on the gaps typically lead to an order-optimal algorithm, as shown in \cite{GarivierETC2016}.

@@

We present the regret bound, which was first stated in \cite{Auer02}, and to explain how Hoeffding's inequality is used we prove it below.
Note that for the rest of the manuscript, we use the \klUCB{} algorithm (and more subtle concentration inequality) whenever we analyze the regret of a newly introduced algorithm (\ie, in Chapters~\ref{chapter:5} and \ref{chapter:6}),
but we think the simpler proof of $\UCB_1$ is enlightening for the reader.
% but we preferred to start with a simpler proof in the single-player stationary case.
%
% This theorem shows that $\UCB_1$ achieves an order-optimal problem-dependent regret, bounded by $\cO(K\log(T)/\Delta^2)$ (if $\Delta \eqdef \min\limits_{k: \Delta_k>0} \Delta$), and the result also has the advantage of being non asymptotic.
This has the advantage of being non asymptotic.

@@

We see in this Theorem~\ref{thm:2:UCBregretBound} that $\UCB_1$ has the advantage of not relying on any prior knowledge of the problem difficulty, of being anytime and computationally simple, for both memory and time aspects.
Moreover, $\UCB_1$ achieves an order-optimal problem-dependent regret upper bound in $\cO((\sum_{k: \Delta_k>0} \frac{1}{\Delta_k})\log(T))$, meaning that it matches the asymptotic regret lower-bound of Lai and Robbins up-to a constant factor \cite{LaiRobbins85} (see the definition above in page \pageref{par:2:def_optimal_orderoptimal}).
%
$\UCB_1$ requires a constant storage capacity for each arm (storing $X_k(t)$ and $N_k(t)$ require two integers or a \texttt{float} and an integer, \ie, $\bigO{1}$), giving a storage capacity of $\bigO{K}$, independently of the horizon $T$.
It also needs to perform some mathematical operations, but a square-root and a logarithm are cheap, thus it costs a $\bigO{1}$ constant time for each arm and time step, thus a time $\bigO{K T}$ for the $T$ time steps.

@@

For Bernoulli distributions, \klUCB{} builds tighter confidence intervals than $\UCB_1$, in the sense that $U_k^{\klUCB}(t) < U_k^{\UCB}(t)$
but with the same probability coverage.
This is justified by Pinkser's inequality, which gives $\kl(x,y) \geq 2(x-y)^2$.
%
The \KLUCB{} and \klUCB{} algorithms achieve finite-time regret logarithmic upper-bounds on their regret, and have been proven to be asymptotically optimal, respectively for one-dimensional exponential families and for binary rewards \cite{Garivier11KL,KLUCBJournal}.
In particular, we highlight that \klUCB{} achieves a finite-time regret upper-bound, that matches the asymptotical lower-bound of Lai and Robbins \cite{LaiRobbins85} for Bernoulli distributed problems, which is our main focus of interest in both Chapters~\ref{chapter:5} and \ref{chapter:6}.

@@

For problems with Bernoulli distributions on the $K$ arms,
the best choice of prior is a Beta distribution, that starts as an initially uniform prior (\ie, $\forall k, \pi_k(0) \eqdef \mathrm{Beta}(1,1)$),
and maintains a distribution $\pi_k(t) \eqdef \mathrm{Beta}(X_k(t)+1,N_k(t)-X_k(t)+1)$,
as $X_k(t)$ counts the number of successes observed on arm $k$, and $N_k(t)-X_k(t)$ counts the number of failures.
%
We detail the Thompson sampling algorithm in this case, below in Algorithm~\ref{algo:2:ThompsonSampling}, as it corresponds to the case used in Chapter~\ref{chapter:4} in Section~\ref{sub:41:numericalResults}.
%
In this setting, as well as for one-dimensional exponential families,
% for Gaussian distributions,
Thompson sampling was proven to achieve finite-time and asymptotical optimal problem-dependent bounds \cite{Kaufmann12Thompson,AgrawalGoyal11}.

@@

Bayes-UCB from \cite{Kaufmann12BUCB} also uses the Bayesian point of view, updated like for Thompson sampling, but used in a different way.
Instead of sampling from the posteriors and playing the arms with maximum sample, Bayes-UCB computes the $(1-1/(t (\log(t))^5)$ quantile of each arm, at time $t$, and plays the arm with the largest quantile.
It was also proven to achieve an asymptotical optimal problem-dependent bound for binary (Bernoulli) bandits \cite{Kaufmann12BUCB}.
%
Bayes-UCB is comparable to Thompson sampling in terms of memory complexity, but it can be slower, as computing a quantile is generally more costly than just sampling from a distribution.
% a Beta prior and posterior, initially uniform (\ie, $\mathrm{Beta}(1,1)$) (\texttt{BayesUCB}).
% But in particular for Bernoulli distributed arms and for Beta posteriors and priors, Bayes-UCB is only about twice as slow as Thompson sampling.

@@

\textbf{BESA} (Best Empirical Sampled Average) was introduced in \cite{Baransi2014}, in order to fix a simple observation:
if arm $1$ has $100$ samples and arm $2$ has $50$ samples, it should make more sense to only use $50$ samples of arm $1$ to compute and compare their empirical means.
Based on this idea, the BESA algorithm for $K=2$ arms is quite simple: at each time step, if arms $1$ and $2$ have respectively $n_1$ and $n_2$ samples, and if $n_1>n_2$ (for instance), first it sub-samples $n_2$ observations from $n_1$, then it computes the mean of these observations, and play the arm maximizing this mean (now that there is the same number of observations for both arms, it makes more sense to compare these quantities).
It was analyzed partially, for the two-armed case it was proven to be asymptotically optimal for one-dimensional exponential families, and the authors illustrated that it can be very efficient empirically.
However, analyzing it in the generic case was found to be difficult, and it is still an open problem.
% We implemented it using a binary tournament implement in a naive but efficient way (not using recursive functions, \ie, using the \texttt{BESA} class with \texttt{''non\_recursive''=True}).
% By using such binary tournament, the extension $K>2$ arms is costly in terms of both time and memory, as illustrated below, and suffers from an exponential blow-up when $K$ increases.

@@

More recently, the \klUCB-switch from \cite{GarivierHadiji2018} is the first algorithm to be proven to
enjoy simultaneously a distribution-free regret bound of optimal order $\sqrt{KT}$, and a distribution-dependent regret bound of optimal order as well, by matching the $C_I \ln(T)$ asymptotic lower-bound by \cite{LaiRobbins85}.
It is an index policy that uses modified versions of the indexes of \klUCB$^+$ for arms that are not sampled much (see \cite{Honda2019} for more details), and of MOSS$^+$ for the other arms.
If the two indexes
% $U^{\KLUCB+}_k(t)$ and $U^{\mathrm{MOSS}+}_k(t)$
are defined by
$U^{\mathrm{MOSS}+}_k(t) \eqdef \frac{X_k(t)}{N_k(t)} + \max\left(0, \sqrt{ \log\left(T / (K N_k(t))\right) / N_k(t)} \right)$
and
$U^{\KLUCB+}_k(t) \eqdef \sup\limits_{q \in [0,1]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \log(T / (K N_k(t))) / N_k(t) \right\}$,
and if $f(T, K) = \lfloor (T/K)^{\gamma}\rfloor$ for $\gamma=1/5$,
then \klUCB-switch uses the index defined as follows:
\begin{align*}
    U^{\klUCB\text{-}\mathrm{switch}}_k(t) & \eqdef \begin{cases}
        U^{\KLUCB+}_k(t) & \text{if } N_k(t) \leq f(T, K), \\
        U^{\mathrm{MOSS}+}_k(t) & \text{if } N_k(t) > f(T, K).
    \end{cases}.
\end{align*}
%

@@

\paragraph{Summary of this review of MAB algorithms.}
%
We gave an overview of the most well known families of algorithms for multi-armed bandits,
focussing on algorithms designed and analyzed for the single-player stationary and stochastic MAB model.
Except the naive and simple strategies given as introductory examples, most of the policies presented above are order-optimal for at least one of the model of Bernoulli distributed problems, bounded rewards or one-dimensional exponential families,
and some of them are known to be optimal in different settings.
%
To clearly understand the difference in their empirical behaviors, we illustrated the performance of two naive and two simple strategies against three efficient ones, on the example of Bernoulli-distributed problem used for the online demonstration from Section~\ref{sec:2:notations} (see Figures~\ref{fig:2:example_of_a_5_arm_bandit_problem} and \ref{fig:2:example_of_a_5_arm_bandit_problem__step100}).
This also serves as a first example of numerical simulation performed with our library SMPyBandits, before the large-scale experiments of Section~\ref{sec:3:reviewSPAlgorithms} below.
%
In the rest of this thesis, we mainly use the $\UCB_1$ or Thompson sampling algorithms when simplicity and realistic deployment is favored (in Chapter~\ref{chapter:4}), and the \klUCB{} algorithm when we give mathematical developments (in Chapters~\ref{chapter:5} and \ref{chapter:6}).

### 2019-07-06 2-Chapters/3-Chapter/chapter3.tex
Not sure if I say enough about the lower-bounds...

\abstractStartChapter{}%
%
SMPyBandits is a Python package I have developed since the beginning of my PhD.
It is designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
It is the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models.
%
This chapter details the organization of the library, what it implements in terms of arm distributions, models, algorithms and visualizations.
Then we use it to perform a numerical comparisons of the main state-of-the-art single-player MAB algorithms, and a study to compare time and memory costs of the main algorithms.

@@

Since December $2016$ I worked regularly on the SMPyBandits library, which is a Python package  designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
%
This library is by far the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models.
It aims at being exhaustive, simple to use and maintain, and has a clean and well documented codebase, and it uses the popular open-source Python language.
It allows fast prototyping of simulations, with an easy configuration system and command-line options to customize experiments.
Experiment results are saved in an optimized binary format (HDF5) as well as high quality plots of useful visualizations.
%
More than two years of active development have shown how easy it can be to add new algorithms, new arm distributions, and new bandit models (\eg, Markov or non-stationary).
It is hosted on GitHub, and uses the state-of-the-art development technologies, by using two online services of continuous integration to run automated tests and to build its online documentation.

>>>>

\abstractStartChapter{}%
%
SMPyBandits is a Python package I have developed since the beginning of my PhD.
It is designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
To the best of the authors' knowledge, it is the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models.
%
This chapter details the organization of the library, what it implements in terms of arm distributions, models, algorithms and visualizations.
Then we use it to perform a numerical comparisons of the main state-of-the-art single-player MAB algorithms, and a study to compare time and memory costs of the main algorithms.

@@

Since December $2016$ I worked regularly on the SMPyBandits library, which is a Python package  designed to allow easy and fast numerical simulations on single-player and multi-players Multi-Armed Bandits (MAB) algorithms.
%
This library is by far the most exhaustive open-source implementation of state-of-the-art algorithms and different kinds of MAB models, according to the authors' knowledge.
It aims at being simple to use and maintain, and has a clean and well documented codebase, and it uses the popular open-source Python language.
It allows fast prototyping of simulations, with an easy configuration system and command-line options to customize experiments.
Experiment results are saved in an optimized binary format (HDF5) as well as high quality plots of useful visualizations.
%
More than two years of active development have shown how easy it can be to add new algorithms, new arm distributions, and new bandit models (\eg, Markov or non-stationary).
It is hosted on GitHub, and uses the state-of-the-art development technologies, by using two online services of continuous integration to run automated tests and to build its online documentation.
